The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-03-19 manuscript/manuscript.Rnw
add missing 'and' and left to right to manuscript
clarify what the housekeeping core placenta figure contains

\begin{figure}[tbh]
<<core.placenta.transcriptome,fig.width=19/2.5,fig.height=6,out.width="190mm",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species) & housekeeping==FALSE,]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]
cptl <- cptl[,median_FPKM:=median(FPKM),by=human_name]
cptl <- cptl[,facet:="high"]
cptl <- cptl[median_FPKM < median(median_FPKM),facet:="low"]
print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
                    labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual(NULL,##"Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      expand_limits(y=c(10^4)) + 
      theme(text=element_text(size=7.5),
            legend.position=c(0.99,0.01),
            axis.text.y=element_text(face="italic",size=7.5),
            legend.margin=margin(0,0,0,0,unit="mm"),
            legend.spacing=unit(0,"mm"),
            legend.justification=c(1,0),
            plot.margin=unit(c(0,0,0,0),"mm"),
            strip.background = element_blank(),
            strip.text.x = element_blank()
            )
      + facet_wrap(~facet,scales="free")
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  ortholog pairs of \human genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if an ortholog pair of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == 'All 13 Eutherians',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to identify housekeeping genes is to use
the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental
\cref{tab:tissue_specificity_index_of_placenta_core_genes} lists the
tissue specificity index for all housekeeping and non-housekeeping
genes; housekeeping genes expressed in all placentas have a lesser
median tissue specificity index than non-housekeeping genes
(Supplemental \cref{fig:tissue_specificity_index_with_housekeeping}).
We also note that housekeeping genes have been defined in \human, and
may not be housekeeping genes in other species; generating tissue
expression atlases and examining the evolution of expression in
non-model organisms is a promising area of future research.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \ac{ANXA1}, \ac{ANXA2},
\ac{S100A10}, \ac{S100A11}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A4, S100A6, S100A10, and S100A11
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\ac{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads}.

>>>>

\begin{figure}[tbh]
<<core.placenta.transcriptome,fig.width=19/2.5,fig.height=6,out.width="190mm",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species) & housekeeping==FALSE,]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]
cptl <- cptl[,median_FPKM:=median(FPKM),by=human_name]
cptl <- cptl[,facet:="high"]
cptl <- cptl[median_FPKM < median(median_FPKM),facet:="low"]
print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
                    labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual(NULL,##"Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      expand_limits(y=c(10^4)) + 
      theme(text=element_text(size=7.5),
            legend.position=c(0.99,0.01),
            axis.text.y=element_text(face="italic",size=7.5),
            legend.margin=margin(0,0,0,0,unit="mm"),
            legend.spacing=unit(0,"mm"),
            legend.justification=c(1,0),
            plot.margin=unit(c(0,0,0,0),"mm"),
            strip.background = element_blank(),
            strip.text.x = element_blank()
            )
      + facet_wrap(~facet,scales="free")
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  ortholog pairs of \human genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom, left to right) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if an ortholog pair of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet the FPKM criterion and are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == 'All 13 Eutherians',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to identify housekeeping genes is to use
the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental
\cref{tab:tissue_specificity_index_of_placenta_core_genes} lists the
tissue specificity index for all housekeeping and non-housekeeping
genes; housekeeping genes expressed in all placentas have a lesser
median tissue specificity index than non-housekeeping genes
(Supplemental \cref{fig:tissue_specificity_index_with_housekeeping}).
We also note that housekeeping genes have been defined in \human, and
may not be housekeeping genes in other species; generating tissue
expression atlases and examining the evolution of expression in
non-model organisms is a promising area of future research.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \ac{ANXA1}, \ac{ANXA2},
\ac{S100A10}, and \ac{S100A11}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A4, S100A6, S100A10, and S100A11
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\ac{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads}.
